---
output:
  html_document:
    variant: markdown_github
    keep_md: true
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  fig.retina = 2,
  out.width = "75%",
  dpi = 96
)
knit_hooks$set(pngquant = hook_pngquant)
```

# fitHeavyTail
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/fitHeavyTail)](https://CRAN.R-project.org/package=fitHeavyTail)
[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/fitHeavyTail)](https://CRAN.R-project.org/package=fitHeavyTail)
[![CRAN Downloads Total](https://cranlogs.r-pkg.org/badges/grand-total/fitHeavyTail?color=brightgreen)](https://CRAN.R-project.org/package=fitHeavyTail)

Robust estimation methods for the mean vector, scatter matrix,
    and covariance matrix (if it exists) from data (possibly containing NAs) 
    under multivariate heavy-tailed distributions such as angular Gaussian 
    (via Tyler's method), Cauchy, and Student's t distributions. Additionally, 
    a factor model structure can be specified for the covariance matrix. The
    latest revision also includes the multivariate skewed t distribution.
    

## Installation
The package can be installed from [CRAN](https://CRAN.R-project.org/package=fitHeavyTail) or [GitHub](https://github.com/convexfi/fitHeavyTail):
```{r, eval=FALSE}
# install stable version from CRAN
install.packages("fitHeavyTail")

# install development version from GitHub
devtools::install_github("convexfi/fitHeavyTail")
```

To get help:
```{r, eval=FALSE}
library(fitHeavyTail)
help(package = "fitHeavyTail")
?fit_mvt
```

To cite [`fitHeavyTail`](https://CRAN.R-project.org/package=fitHeavyTail) in publications:
```{r, eval=FALSE}
citation("fitHeavyTail")
```


## Quick Start
To illustrate the simple usage of the package [`fitHeavyTail`](https://CRAN.R-project.org/package=fitHeavyTail), let's start by generating some multivariate data under a Student's $t$ distribution with significant heavy tails (degrees of freedom $\nu=4$):
```{r}
library(mvtnorm)  # package for multivariate t distribution
N <- 10   # number of variables
T <- 80   # number of observations
nu <- 4   # degrees of freedom for heavy tails

set.seed(42)
mu <- rep(0, N)
U <- t(rmvnorm(n = round(0.3*N), sigma = 0.1*diag(N)))
Sigma_cov <- U %*% t(U) + diag(N)  # covariance matrix with factor model structure
Sigma_scatter <- (nu-2)/nu * Sigma_cov
X <- rmvt(n = T, delta = mu, sigma = Sigma_scatter, df = nu)  # generate data
```

We can first estimate the mean vector and covariance matrix via the traditional sample estimates (i.e., sample mean and sample covariance matrix):
```{r}
mu_sm     <- colMeans(X)
Sigma_scm <- cov(X)
```

Then we can compute the robust estimates via the package [`fitHeavyTail`](https://CRAN.R-project.org/package=fitHeavyTail):
```{r}
library(fitHeavyTail)
fitted <- fit_mvt(X)
```

We can now compute the estimation errors and see the significant improvement:
```{r}
sum((mu_sm     - mu)^2)
sum((fitted$mu - mu)^2)

sum((Sigma_scm  - Sigma_cov)^2)
sum((fitted$cov - Sigma_cov)^2)
```

```{r, eval=FALSE, echo=FALSE}
# fitting with factor model
fitted_3factors <- fit_mvt(X, factors = 3)
sum((fitted_3factors$mu - mu)^2)
sum((fitted_3factors$cov - Sigma_cov)^2)
```

To get a visual idea of the robustness, we can plot the shapes of the covariance matrices (true and estimated ones) on two dimensions. Observe how the heavy-tailed estimation follows the true one more closely than the sample covariance matrix:

```{r scatter-plots, echo=FALSE, message=FALSE, fig.width=10, fig.height=6, out.width="80%"}
# fig.cap="Sample covariance matrix vs robust estimator."
library(mvtnorm)
library(ellipse)
library(ggplot2)

N <- 2
T <- 50
nu <- 4
mu <- rep(0, N)
Sigma <- matrix(c(0.00125, 0.00112, 0.00112, 0.00125), N, N)

set.seed(42)
X <- rmvt(n = 200, delta = mu, sigma = (nu-2)/nu * Sigma, df = nu)
X_ <- X[1:T, ]
Sigma_scm       <- cov(X_)
Sigma_heavytail <- fit_mvt(X_, optimize_mu = FALSE, nu = 4, scale_covmat = TRUE)$cov
colnames(Sigma_heavytail) <- rownames(Sigma_heavytail) <- NULL

# scatter plot
ggplot(data.frame(x = X[, 1], y = X[, 2]), aes(x, y)) +
  geom_point(alpha = 1, size = 0.8) +
  geom_path(data = data.frame(ellipse(Sigma)),           aes(x, y, col = "1"), linewidth = 1) +
  geom_path(data = data.frame(ellipse(Sigma_scm)),       aes(x, y, col = "2"), linewidth = 1) +
  geom_path(data = data.frame(ellipse(Sigma_heavytail)), aes(x, y, col = "3"), linewidth = 1) +
  coord_cartesian(xlim = c(-0.15, 0.15), ylim = c(-0.15, 0.15)) +
  scale_color_manual(name = "ellipses",
                     values = c("1" = "black", "2" = "#c0392b", "3" = "#2980b9"),
                     labels = c("true", "SCM estimation", "heavy-tailed estimation")) +
  labs(title = "Shape of covariance matrices", x = NULL, y = NULL)
```




## Documentation
For more detailed information, please check the
[vignette](https://CRAN.R-project.org/package=fitHeavyTail/vignettes/CovarianceEstimationHeavyTail.html).



## Links
Package: [CRAN](https://CRAN.R-project.org/package=fitHeavyTail) and [GitHub](https://github.com/convexfi/fitHeavyTail).

README file: [GitHub-readme](https://github.com/convexfi/fitHeavyTail/blob/master/README.md).

Vignette: [CRAN-vignette](https://CRAN.R-project.org/package=fitHeavyTail/vignettes/CovarianceEstimationHeavyTail.html) and [GitHub-vignette](https://htmlpreview.github.io/?https://github.com/convexfi/fitHeavyTail/blob/master/vignettes/CovarianceEstimationHeavyTail.html).

