---
always_allow_html: yes
output:
  md_document:
    variant: markdown_github
  html_document:
    keep_md: yes
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  fig.retina = 2,
  out.width = "75%",
  dpi = 96
)
knit_hooks$set(pngquant = hook_pngquant)
```



# fitHeavyTail
<!---
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/fitHeavyTail)](https://CRAN.R-project.org/package=fitHeavyTail)
[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/fitHeavyTail)](https://CRAN.R-project.org/package=fitHeavyTail)
[![CRAN Downloads Total](https://cranlogs.r-pkg.org/badges/grand-total/fitHeavyTail?color=brightgreen)](https://CRAN.R-project.org/package=fitHeavyTail)
--->

Robust estimation methods for the mean vector and covariance matrix 
from data (possibly containing NAs) under multivariate heavy-tailed 
distributions such as angular Gaussian, Cauchy, and Student's $t$. Additionally,
a factor model structure can be specified for the covariance matrix.
    

## Installation
The package can be installed from [CRAN](https://CRAN.R-project.org/package=fitHeavyTail) or [GitHub](https://github.com/dppalomar/fitHeavyTail):
```{r, eval=FALSE}
# install stable version from CRAN
install.packages("fitHeavyTail")

# install development version from GitHub
devtools::install_github("dppalomar/fitHeavyTail")
```

To get help:
```{r, eval = FALSE}
library(fitHeavyTail)
help(package = "fitHeavyTail")
?fit_mvt
```

To cite `fitHeavyTail` in publications:
```{r, eval = FALSE}
citation("fitHeavyTail")
```


## Quick Start
To illustrate the simple usage of the package `fitHeavyTail`, let's start by generating some multivariate data under a Student's $t$ distribution with significant heavy tails ($\nu=4$):
```{r}
library(mvtnorm)  # package for multivariate t distribution
N <- 10   # number of variables
T <- 80   # number of observations
nu <- 4   # degrees of freedom for tail heavyness

set.seed(42)
mu <- rep(0, N)
U <- t(rmvnorm(n = round(0.3*N), sigma = 0.1*diag(N)))
Sigma <- U %*% t(U) + diag(N)  # covariance matrix with factor model structure
Sigma_scatter <- (nu-2)/nu * Sigma
X <- rmvt(n = T, delta = mu, sigma = Sigma_scatter, df = nu)  # generate data
```

We can first estimate the mean vector and covariance matrix via the traditional sample estimates (i.e., sample mean and sample covariance matrix):
```{r}
mu_sm <- colMeans(X)
Sigma_scm <- cov(X)
```

Then we can compute the robust estimates via the package `fitHeavyTail`:
```{r}
library(fitHeavyTail)
fitted <- fit_mvt(X)
```

We can now compute the estimation errors and see the big improvement:
```{r}
sum((mu_sm - mu)^2)
sum((fitted$mu - mu)^2)

sum((Sigma_scm - Sigma)^2)
sum((fitted$cov - Sigma)^2)
```

To get a visual idea of the robustness, we can plot the shapes of the covariance matrices (true and estimated ones) projected on two dimensions. Observe how the heavy-tailed estimation follows the true one more closely than the sample covariance matrix:
```{r ellipses, echo=FALSE, message=FALSE, fig.width = 9, fig.height = 5, out.width = "80%"}
library(ggplot2)
library(ellipse)

i1 <- 7; i2 <- 5
colnames(fitted$cov) <- rownames(fitted$cov) <- NULL
ggplot(data.frame(x = X[, i1], y = X[, i2]), aes(x, y)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_path(data = data.frame(ellipse(Sigma[c(i1, i2), c(i1, i2)])), 
            aes(x, y, col = "true"), lwd = 1) +
  geom_path(data = data.frame(ellipse(Sigma_scm[c(i1, i2), c(i1, i2)], centre = mu_sm[c(i1, i2)])), 
            aes(x, y, col = "SCM estimation"), lwd = 1) +
  geom_path(data = data.frame(ellipse(fitted$cov[c(i1, i2), c(i1, i2)], centre = fitted$mu[c(i1, i2)])), 
            aes(x, y, col = "heavy-tailed estimation"), lwd = 1) +
  scale_colour_manual(name = "ellipses",
                      values = c("true" = "black", "SCM estimation" = "red", "heavy-tailed estimation" = "blue")) +
  guides(color = guide_legend(reverse = TRUE)) +
  labs(title = "Data points and shapes of covariance matrices", x = NULL, y = NULL)
```


<!---
## Documentation
For more detailed information, please check the
[vignette](https://CRAN.R-project.org/package=fitHeavyTail/vignettes/CovarianceEstimationHeavyTail.html).
--->


## Links
Package: [CRAN](https://CRAN.R-project.org/package=fitHeavyTail) and [GitHub](https://github.com/dppalomar/fitHeavyTail).

README file: [GitHub-readme](https://github.com/dppalomar/fitHeavyTail/blob/master/README.md).
<!---
Vignette: [CRAN-vignette](https://CRAN.R-project.org/package=fitHeavyTail/vignettes/CovarianceEstimationHeavyTail.html).
--->
