---
title: "Estimation of nu"
author: |
  | Daniel P. Palomar
  | The Hong Kong University of Science and Technology (HKUST)
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: vignette
    toc: yes
    toc_depth: 2
  html_document:
    theme: flatly
    highlight: pygments  
    toc: yes
    toc_depth: 2
params:
  N_realiz: 50
csl: ieee.csl
bibliography: refs.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.retina = 2,
  out.width = "85%",
  dpi = 96,
  pngquant = "--speed=1"
)
knitr_in_progress <- isTRUE(getOption('knitr.in.progress'))
knit_hooks$set(pngquant = hook_pngquant)
# rmarkdown::render("vignettes/nu_estimation.Rmd", "prettydoc::html_pretty", params = list(N_realiz = 500))
```

-----------


Let's define the parameters for the generation of the multivariate Student's $t$ distribution:
```{r}
library(mvtnorm)  # package for multivariate t distribution
set.seed(137)

N <- 10   # number of variables
T <- 80   # number of observations
nu <- 4   # degrees of freedom for tail heavyness
mu <- rep(0, N)
U <- t(rmvnorm(n = round(0.3*N), sigma = 0.1*diag(N)))
Sigma <- U %*% t(U) + diag(N)  # covariance matrix with factor model structure
Sigma_scatter <- (nu-2)/nu * Sigma
```


To warm up, let's consider the following estimators for $\nu$:

- based on the kurtosis directly
- based on EM with initial value of 4
- based on EM with initial value from kurtosis

```{r}
library(fitHeavyTail)

X <- rmvt(n = T, delta = mu, sigma = Sigma_scatter, df = nu)  # generate data
nu_kurtosis  <- fit_mvt(X, nu = fitHeavyTail:::est_nu_kurtosis(X))$nu
nu0_4        <- fit_mvt(X, initial = list("nu" = 4))$nu
nu0_kurtosis <- fit_mvt(X, initial = list("nu" = fitHeavyTail:::est_nu_kurtosis(X)))$nu
rbind("true" = nu, nu_kurtosis, nu0_4, nu0_kurtosis)
```


We now run `r params$N_realiz` Monte Carlo simulations.
```{r, echo=FALSE}
eval_single <- function(X) {
  nu <- list()
  nu["nu = kurt"]  <- fit_mvt(X, nu = fitHeavyTail:::est_nu_kurtosis(X))$nu
  nu["nu0 = 4"]    <- fit_mvt(X, initial = list("nu" = 4))$nu
  nu["nu0 = kurt"] <- fit_mvt(X, initial = list("nu" = fitHeavyTail:::est_nu_kurtosis(X)))$nu
  return(nu)
}

T_sweep <- round(seq(from = 25, to = 50, by = 5))
if (!knitr_in_progress) pbar <- txtProgressBar(min = it <- 0, max = length(T_sweep), style = 3)
nu_all_T <- list()
for(T in T_sweep) {
  if (!knitr_in_progress) setTxtProgressBar(pbar, it <- it + 1)
  
  # first, generate random heavy-tailed data sequentially for reproducibility
  X <- replicate(params$N_realiz, rmvt(n = T, delta = mu, sigma = Sigma_scatter, df = nu), simplify = FALSE)
  names(X) <- paste0("realiz ", 1:params$N_realiz)

  # then, run estimations for all realizations
  nu_all_T <- c(nu_all_T, list(lapply(X, eval_single)))
}
names(nu_all_T) <- paste0("T = ", T_sweep)
```

The results can be see in the following boxplots:
```{r, echo=FALSE, fig.width = 9, fig.height = 5, out.width = "100%"}
library(ggplot2)
library(reshape2)

nu_all_T_molten <- melt(nu_all_T)
names(nu_all_T_molten) <- c("value", "method", "realization", "T")
nu_all_T_molten$T <- factor(nu_all_T_molten$T, levels = names(nu_all_T))

ggplot(nu_all_T_molten, aes(x = T, y = value, fill = method)) +
  geom_boxplot(show.legend = FALSE) +  #show.legend = FALSE
  geom_point(size = 0.5, alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~ method) +
  scale_x_discrete(labels = T_sweep) +
  labs(title = sprintf("Estimation of nu (nu = %d, N = %d)", nu, N), x = "T", y = NULL)
```

```{r, echo=FALSE, eval=FALSE, fig.width = 9, fig.height = 5, out.width = "100%"}
ggplot(nu_all_T_molten, aes(x = method, y = value, fill = method)) +
  geom_boxplot() + 
  geom_point(size = 0.5, alpha = 0.8, show.legend = FALSE) +
  facet_grid(~ T) +
  scale_x_discrete(labels = element_blank()) +  #theme(axis.text.x = element_blank()) +
  labs(title = sprintf("Estimation of nu (nu = %d, N = %d)", nu, N), x = NULL, y = NULL)
```

We can see that the EM algorithm generally gives an improvement with respect to the direct estimation based on the marginal kurtosis, but it is subject to unacceptable outliers. Perhaps with the usage of RMT techniques these outliers can be eliminated?










