---
title: "Comparison of Covariance Matrix Estimation with different treatment on nu"
author: Rui ZHOU
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.retina = 2,
  out.width = "85%",
  dpi = 96,
  pngquant = "--speed=1"
)
knitr_in_progress <- isTRUE(getOption('knitr.in.progress'))
knit_hooks$set(pngquant = hook_pngquant)
# rmarkdown::render("R_buildignore/tricks_on_nu.Rmd", "prettydoc::html_pretty")
```

-----------


# Numerical Comparison with Existing Benchmarks


```{r}
library(mvtnorm)
library(fitHeavyTail)
library(ggplot2)
library(reshape2)

# Comparison with other packages:
#   EstimateMoments_MinskerWei(X) is not good...
#   rrcov::

N <- 20
nu <- 5
mu <- rep(0, N)

set.seed(123)
U <- t(rmvnorm(n = round(1.1*N), sigma = 0.1*diag(N)))
Sigma <- U %*% t(U) + diag(N)
Sigma_scale <- (nu-2)/nu * Sigma
qplot(eigen(Sigma)$values, geom = "histogram", xlab = "eigenvalues", fill = I("cyan"), col = I("black"),
      main = "Histogram of eigenvalues of true covariance matrix")
```

```{r, fig.width = 9, fig.height = 5, out.width = "100%", message = FALSE, cache = TRUE}
# compute the MSE of estimated result with the reference one
MSE <- function(est_cov) norm(est_cov - Sigma, "F")^2
eval_single_res <- function(X) {
  c("MLE" = MSE(fit_mvt(X)$cov),
    "MLE (nu = 6)" = MSE(fit_mvt(X, nu = 6)$cov),
    "MLE (nu from kurtosis)" = MSE(fit_mvt(X, nu_regcoef = 1e10)$cov),
    "MLE (nu_target = 6)" = MSE(fit_mvt(X, nu_target = 6, nu_regcoef = 5e-2)$cov),
    "MLE (nu_target from kurtosis)" = MSE(fit_mvt(X, nu_regcoef = 5e-2)$cov),
    "MLE + correct" = MSE(fit_mvt(X, scale_correct = TRUE)$cov),
    "MLE (nu = 6) + correct" = MSE(fit_mvt(X, nu = 6, scale_correct = TRUE)$cov),
    "MLE (nu from kurtosis) + correct" = MSE(fit_mvt(X, nu_regcoef = 1e10, scale_correct = TRUE)$cov),
    "MLE (nu_target = 6) + correct" = MSE(fit_mvt(X, nu_target = 6, nu_regcoef = 5e-2, scale_correct = TRUE)$cov),
    "MLE (nu_target from kurtosis) + correct" = MSE(fit_mvt(X, nu_regcoef = 5e-2, scale_correct = TRUE)$cov))
}
N_realiz <- 200  # multiple realizations for averaging
T_sweep <- round(seq(from = ceiling(1.5*N), to = 5*N, length.out = 12))

if (!knitr_in_progress) pbar <- txtProgressBar(min = it<-0, max = length(T_sweep), style=3)
MSE_all_T <- NULL
for(T in T_sweep) {
  if (!knitr_in_progress) setTxtProgressBar(pbar, it<-it+1)
  res <- sapply(1:N_realiz, function(idx) {
    X <- rmvt(n = T, delta = mu, sigma = Sigma_scale, df = nu)
    eval_single_res(X)
  })
  MSE_all_T <- rbind(MSE_all_T, apply(res, 1, mean))
}


# MSE plots
rownames(MSE_all_T) <- T_sweep
ggplot(melt(MSE_all_T), aes(x = Var1, y = value, col = Var2, shape = Var2)) +
  geom_line() + geom_point() + coord_cartesian(ylim = c(0, 250)) +
  theme(legend.title = element_blank()) +
  ggtitle(bquote("MSE of covariance matrix estimation for heavy-tailed data (" * N == .(N) * "," ~ nu == .(nu)* ")")) +
  xlab("T") + ylab("MSE")

```

Observations:

- Set the $\nu$ to be the estimator from kurtosis method seems to give the best estimation performance for covariance matrix.
- Set the $\nu$ to be fixed $6$ provides also very good estimation performance, while alittle worse than the kurtosis estimator.
- If the implementation for re-scaling is correct, then it seems that the correct operating does not work, except the first stage estimation is really terriable.



\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent
